General purpose gene annotation using deep learning

### 2020-01-09 paper/main.tex
datasets update to 50k default for non-invertebrates
post processing roughly

Once a set of training genomes was selected, we generated a validation set from it by splitting of 20\% of the FASTA sequences above and below the N90 of each genome, ensuring a proper distribution of large and short sequences in both the training and validation set. The validation set was only used to roughly judge the performance of our models during training. Evaluation of the annotations on all training and validation genomes was done regularly, as we ultimately only cared about the performance on fully unknown genomes.

@@

\subsection{Model Architecture and Usage}
	We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

The model was trained with 10 bases of genomic sequence as input during each time step and produces individual predictions for each of those 10 bases simultaneously. This grouping enabled us to train with far longer sequences than normally done.

- how we train and validate the model
- input of longer sequences
- overlapping
- ensembling

>>>>

Once a set of training genomes was selected, we generated a validation set from it by splitting of 20\% of the FASTA sequences above and below the N90 of each genome, ensuring a proper distribution of large and short sequences in both the training and validation set and a split on the chromosome level. The validation set was only used to roughly judge the performance of our models during training. Evaluation of the annotations on all training and validation genomes was done regularly, as we ultimately only cared about the performance on for the model fully unknown genomes.

@@

\subsection{Model Architecture and Usage}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{Tab:02} or in the Helixer source code repository.

The model was trained with 10 bases of genomic sequence as input during each time step and produces individual predictions for each of those 10 bases simultaneously. This grouping enabled us to train with far longer sequences than usual. A class wise evaluation including the calculation of the Genic F1 score is performed for the validation set after every epoch. We use the model with the best Genic F1 score unless the improvement to the next best one was less than $0.01$ apart in order to prevent overfitting.

We also used multiple techniques to improve the prediction quality after the training was done. One very effective way was to input longer sequences during inference than during training. The sequences were up to 10 times larger, depending on species group and assembly quality. We also reconstructed the final predictions of a single model from overlapping predictions that were cut to a core sequence. This was done to drastically reduce a typical bias of the models towards the beginning and end of each sequence and also to provide the model with multiple different starting points. The different overlapping sequences were recombined by averaging the softmax values of each base. This was also done when we finally ensembled the whole output of different models on the softmax value level from (8?) different models. For more details on these post-processing techniques see Suplementary Material (?).
